---
title: "SDS 315 Homework 7 Response"
author: "Grace Lefferts (gel766)"
date: "2025-04-07"
output: pdf_document
urlcolor: blue
---

[GitHub Link](https://github.com/graceleff05/SDS315_HW7.git)

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=80), echo=FALSE)
```

```{r, message=FALSE}
library(tidyverse)
library(kableExtra)
library(mosaic)
library(dplyr)
library(MatchIt)
armfold = read.csv('armfold.csv')
turnout = read.csv('turnout.csv')
```

## **Problem 1: Armfolding**

### Part A: # of Male/Female Students and Proportions who Folded Left Arm on Top

-   **Number of male students:** `r sum(armfold$Sex == 'Male')`

-   **Number of female students:** `r sum(armfold$Sex == 'Female')`

```{r}
# Calculate proportion of males & females who folded left arm on top
armfold_props = prop(LonR_fold ~ Sex, data=armfold)
```

-   **Sample proportion of males who folded their left arm on top:** `r round(armfold_props[2:2], 3)`

-   **Sample proportion of females who folded their left arm on top:** `r round(armfold_props[1:1], 3)`

### Part B: Difference in Proportions

```{r}
# Calculating observed difference in proportion of males & proportion of females who fold their left arm on top
observed_diff = round(armfold_props[2:2] - armfold_props[1:1], 3)
```

-   **Observed difference in proportions (males - females):** `r observed_diff`

### Part C: 95% Confidence Interval

```{r, results = 'hide'}
# Calculating two 95% confidence intervals for difference in proportions
prop.test(LonR_fold ~ Sex, data=armfold)
lower = observed_diff - 1.96*.067
upper = observed_diff + 1.96*.067
```

-   **Results from prop.test() function:**
    -   CI = [-0.093, 0.190]
-   **Standard error:** 0.067

```{r, echo=TRUE, results='hide'}
sqrt((0.472)*(1-0.472)/106 + (0.423)*(1-0.423)/111)
```

-   **z\* value:** 1.96; 95% of the time, the result will be within 1.96 standard deviations of the observed difference in proportions
-   **Result from hand calculations:** [-0.083, 0.179]
    -   **Lower bound:** 0.048 - 1.96\*0.067 = -0.083
    -   **Upper bound:** 0.048 + 1.96\*0.067 = 0.179

### Part D: Interpretation

If we were to take many, many repeated random samples from the population of Australian university students and calculate a confidence interval for each, we would expect 95% of these confidence intervals to contain the true population sex difference in proportions.

### Part E: Standard Error

The standard error calculated above represents the standard deviation of the sample, which is a measure of variation of the sample sex difference in arm-folding.

### Part F: Sampling Distribution

In this context, a sampling distribution would consist of the differences in proportions for many samples of men and women and their arm crossing habits (Which changes from sample to sample), but the sampling size (number of men and women) would be the same for each sample.

### Part G: Central Limit Theorem

The Central Limit Theorem justifies using a normal distribution to approximate the sampling distribution of the difference in sample proportions in that the sampling distributions are almost always normally distributed, as long as the sample size is large enough.

### Part H: CI Containing 0

Given a 95% confidence interval for the difference of proportions of [-0.01, 0.30], because 0 is contained in the interval, it can be said that there is no significant sex difference in armfolding. There could be a difference, but we cannot claim that there is based on this interval.

### Part I: Repetition of Experiment

Repeating this experiment many times with different random samples of university students would create different confidence intervals because the people in the samples would not be identical. However, we would expect that 95% of the confidence intervals would contain the true population difference in proportions.

\newpage

## **Problem 2: Get out the vote**

### Part A: How much more likely are GOTV call recipients to have voted in 1998?

```{r, results = 'hide'}
# Calculate proportion of 1998 voters who did/did not receive GOTV calls
GOTV98_props = prop(voted1998 ~ GOTV_call, data=turnout)

# Calculate large-sample 95% confidence interval for difference in above proportions
prop.test(voted1998 ~ GOTV_call, data=turnout)
```

-   **Sample proportion of those who received a GOTV call who voted in 1998:** `r round(GOTV98_props[2:2], 3)`
-   **Sample proportion of those who did not receive a GOTV call who voted in 1998:** `r round(GOTV98_props[1:1], 3)`
-   **Confidence interval (GOTV call voters - nonGOTV call voters):**
    -   CI = [0.141, 0.266]

### Part B
 
#### Analyzing effect of voted1996 variable on GOTV call reception and voting in 1998
```{r}
# Proportions for GOVT_call and voted1996
GOTV96_props = prop(voted1996 ~ GOTV_call, data=turnout)
```

```{r, results='hide'}
#CI for difference in proportions of those who did/did not receive a GOTV call who voted in 1996 (did not receive call - received call)
prop.test(voted1996 ~ GOTV_call, data=turnout)
```

-   **Sample proportion of those who received a GOTV call who voted in 1996:** `r round(GOTV96_props[2:2], 3)`
-   **Sample proportion of those who did not receive a GOTV call who voted in 1996:** `r round(GOTV96_props[1:1], 3)`
-   **Sample difference in proportions:** `r round(GOTV96_props[2:2]-GOTV96_props[1:1], 3)`
-   **Confidence interval (GOTV call voters - nonGOTV call voters):**
    -   CI = [0.122, 0.241]
-   **Analysis:** The voted1996 variable is a confounding variable in preventing the difference observed in Part A from representing the true causal effect of the GOTV call on the likelihood that a person voted in 1998 because whether someone voted in 1996 influences whether they received a GOTV call and therefore whether they voted in 1998 (since receiving a GOTV call itself influences whether someone voted in 1998). This is illustrated by the above confidence interval. The 95% confidence interval for the difference in proportions of those who received a GOTV call who voted in 1996 and those who did not receive a GOTV call who voted in 1996 suggests that this difference is somewhere between 0.122 and 0.241, or 12.2% and 24.1%. This confidence interval suggests a significant difference between the proportions, thus influencing the relationship between GOTV_call and voted1998 and solidifying its status as a confounding variable.
    
#### Analyzing effect of AGE variable on GOTV call reception and voting in 1998
```{r}
# Grouping turnout by GOTV_call and calculating average age for each group
age_GOTV = turnout %>%
  group_by(GOTV_call) %>%
  summarize(avg_age = mean(AGE))
```

```{r, results='hide'}
# CI for difference in mean age of those who received a GOTV call and those who did not
t.test(AGE ~ GOTV_call, data=turnout)
```

-   **Mean age of those who did not receive a GOTV call:** `r round(age_GOTV[1,2], 3)` years
-   **Mean age of those who received a GOTV call:** `r round(age_GOTV[2,2], 3)` years
-   **Sample difference in mean age:** `r round(age_GOTV[1,2]-age_GOTV[2,2], 3)` years
-   **Confidence interval (mean age of nonGOTV call voters - mean age of GOTV call voters):**
    -   CI = [-11.395, -6.370]
-   **Analysis:**  The AGE variable is a confounding variable in preventing the difference observed in Part A from representing the true causal effect of the GOTV call on the likelihood that a person voted in 1998 because someone’s age influences whether they received a GOTV call and therefore whether they voted in 1998 (since receiving a GOTV call itself influences whether someone voted in 1998). This is illustrated by the above confidence intervals. The 95% confidence interval for the difference in means of those who received a GOTV call and those who did not receive a GOTV call suggests that this difference is somewhere between -11.395 and -6.370 years (mean age of those who did not receive a call – that of those who did). This confidence interval suggests a significant difference between the means, thus influencing the relationship between GOTV_call and voted1998 and solidifying its status as a confounding variable.

#### Analyzing effect of MAJORPTY variable on GOTV call reception and voting in 1998
```{r}
# Proportions for GOVT_call and MAJORPTY
GOTVmajor_props = prop(MAJORPTY ~ GOTV_call, data=turnout)
```

```{r, results='hide'}
#CI for difference in proportions of those who did/did not receive a GOTV call who were registered with a major party (did not receive call - received call)
prop.test(MAJORPTY ~ GOTV_call, data=turnout)
```

-   **Sample proportion of those who received a GOTV call who were registered with a major party:** `r round(GOTVmajor_props[2:2], 3)`
-   **Sample proportion of those who did not receive a GOTV call who were registered with a major party:** `r round(GOTVmajor_props[1:1], 3)`
-   **Sample difference in proportions:** `r round(GOTVmajor_props[2:2]-GOTVmajor_props[1:1], 3)`
-   **Confidence interval (GOTV call voters - nonGOTV call voters):**
    -   CI = [0.004, 0.109]
-   **Analysis:** The MAJORPTY variable is a confounding variable in preventing the difference observed in Part A from representing the true causal effect of the GOTV call on the likelihood that a person voted in 1998 because whether someone is registered with a major party influences whether they received a GOTV call and therefore whether they voted in 1998 (since receiving a GOTV call itself influences whether someone voted in 1998). This is illustrated by the above confidence intervals. The 95% confidence interval for the difference in proportions of those who received a GOTV call who were registered with a major party and those who did not receive a GOTV call who were registered with a major party suggests that this difference is somewhere between 0.004 and 0.109, or 0.4% and 10.9%. This confidence interval suggests a significant difference between the proportions, thus influencing the relationship between GOTV_call and voted1998 and solidifying its status as a confounding variable.

### Part C

```{r}
# Creating a matched dataset to eliminate confounders (voted1996, AGE, MAJORPTY)
turnout_matched = matchit(GOTV_call ~ voted1996 + AGE + MAJORPTY, data = turnout, ratio=5) %>%
  match.data
```

#### Analyzing influence of voted1996 variable under matched data
```{r}
# Proportions for GOVT_call and voted1996
GOTV96_props_matched = prop(voted1996 ~ GOTV_call, data=turnout_matched)
```

```{r, results='hide'}
#CI for difference in proportions of those who did/did not receive a GOTV call who voted in 1996 (did not receive call - received call)
prop.test(voted1996 ~ GOTV_call, data=turnout_matched)
```

-   **Sample proportion of those who received a GOTV call who voted in 1996:** `r round(GOTV96_props_matched[2:2], 3)`
-   **Sample proportion of those who did not receive a GOTV call who voted in 1996:** `r round(GOTV96_props_matched[1:1], 3)`
-   **Sample difference in proportions:** `r round(GOTV96_props_matched[2:2]-GOTV96_props_matched[1:1], 3)`
-   **Confidence interval (GOTV call voters - nonGOTV call voters):**
    -   CI = [-0.062, 0.062]
-   **Analysis:** The voted1996 variable is no longer a confounding variable in preventing the difference in Part A from representing the true causal effect of the GOTV call on the likelihood that a person voted in 1998 because under the matched data, the proportion of those who received a GOTV call who voted in 1996 is the same as the proportion of those who did not receive a GOTV call who voted in 1996 (see above). The 95% confidence interval illustrates that the difference between these proportions is somewhere between -0.062 and 0.062. Because 0 is contained in the interval, it cannot be said that there is a significant difference between the proportions. Thus, voted1996 is no longer a confounding variable.
    
#### Analyzing influence of AGE variable under matched data
```{r}
# Grouping turnout by GOTV_call and calculating average age for each group
age_GOTV_matched = turnout_matched %>%
  group_by(GOTV_call) %>%
  summarize(avg_age = mean(AGE))
```

```{r, results='hide'}
# CI for difference in mean age of those who received a GOTV call and those who did not
t.test(AGE ~ GOTV_call, data=turnout_matched)
```

-   **Mean age of those who did not receive a GOTV call:** `r round(age_GOTV_matched[1,2], 3)` years
-   **Mean age of those who received a GOTV call:** `r round(age_GOTV_matched[2,2], 3)` years
-   **Sample difference in mean age:** `r round(age_GOTV_matched[1,2]-age_GOTV_matched[2,2], 3)` years
-   **Confidence interval (mean age of nonGOTV call voters - mean age of GOTV call voters):**
    -   CI = [-2.760, 2.678]
-   **Analysis:** The AGE variable is no longer a confounding variable in preventing the difference in Part A from representing the true causal effect of the GOTV call on the likelihood that a person voted in 1998 because under the matched data, the mean age of those who did not receive a GOTV call is virtually the same as the mean age of those who received a GOTV call (see above). The 95% confidence interval illustrates that the difference between these means is somewhere between -2.760 and 2.678 years. Because 0 is contained in the interval, it cannot be said that there is a significant difference between the means. Thus, AGE is no longer a confounding variable.

#### Analyzing influence of MAJORPTY variable under matched data
```{r}
# Proportions for GOVT_call and MAJORPTY
GOTVmajor_props_matched = prop(MAJORPTY ~ GOTV_call, data=turnout_matched)
```

```{r, results='hide'}
#CI for difference in proportions of those who did/did not receive a GOTV call who were registered with a major party (did not receive call - received call)
prop.test(MAJORPTY ~ GOTV_call, data=turnout_matched)
```

-   **Sample proportion of those who received a GOTV call who were registered with a major party:** `r round(GOTVmajor_props_matched[2:2], 3)`
-   **Sample proportion of those who did not receive a GOTV call who were registered with a major party:** `r round(GOTVmajor_props_matched[1:1], 3)`
-   **Sample difference in proportions:** `r round(GOTVmajor_props_matched[2:2]-GOTVmajor_props_matched[1:1], 3)`
-   **Confidence interval (GOTV call voters - nonGOTV call voters):**
    -   CI = [-0.062, 0.051]
-   **Analysis:** The MAJORPTY variable is no longer a confounding variable in preventing the difference in Part A from representing the true causal effect of the GOTV call on the likelihood that a person voted in 1998 because under the matched data, the proportion of those who received a GOTV call who were registered with a major party is the same as the proportion of those who did not receive a GOTV call who were registered with a major party (see above). The 95% confidence interval illustrates that the difference between these proportions is somewhere between -0.062 and 0.051. Because 0 is contained in the interval, it cannot be said that there is a significant difference between the proportions. Thus, MAJORPTY is no longer a confounding variable.

#### Analysis from Part A, but with matched data set
```{r, results='hide'}
# Calculate proportion of 1998 voters who did/did not receive GOTV calls
GOTV98_props_matched = prop(voted1998 ~ GOTV_call, data=turnout_matched)

# Calculate large-sample 95% confidence interval for difference in above proportions
prop.test(voted1998 ~ GOTV_call, data=turnout_matched)
```

-   **Sample proportion of those who received a GOTV call who voted in 1998:** `r round(GOTV98_props_matched[2:2], 3)`
-   **Sample proportion of those who did not receive a GOTV call who voted in 1998:** `r round(GOTV98_props_matched[1:1], 3)`
-   **Results from prop.test() function (GOTV call voters - nonGOTV call voters):**
    -   CI = [0.010, 0.147]
-   **Conclusion:** Based on the matched data, it can be concluded with 95% confidence that the difference between the proportion of those who received a GOTV call who voted in 1998 and those who did not receive a GOTV call who voted in 1998 is somewhere between a difference of 0.010 and 0.147, or 1.0% and 14.7%. In other words, receiving a GOTV call increases the likelihood of voting in the 1998 congressional elections.
